Review Paper Draft

### 2020-08-12 main.tex
Section 2: revisions

Population structure is a common phenomenon in genetic data. Varying levels of relatedness are almost always present among genetic samples, even in samples of unrelated individuals and seemingly homogeneous populations. For example, European American \citep{campbell2005demonstrating}, Han Chinese \citep{xu2009genomic, chen2009genetic}, and most recently, cohorts within the UK Biobank data \citep{haworth2019apparent}.

Relatedness describes how genetically similar individuals are, and can refer to both ancient and recent relatedness.  Recent relatedness is characterized by small clusters of high levels of similarity and is indicative of familial or \textbf{cryptic relatedness}. Ancient relatedness, or common ancestry, is generally characterized by large clusters of low levels of similarity. The presence of multiple distinct ancestry groups is known as \textbf{population stratification}.

Pedigree-based methods may be used to explicitly model recent relatedness if familial relationships are known. However, in this review we focus on methods to account for unobserved relatedness. Throughout this paper, the terms \textbf{population structure} and \textbf{structured population} will be used to describe a sample of individuals in which population stratification, cryptic relatedness, or both are present. Population stratification in particular has been of great concern in genetic studies due to its potential to lead to spurious associations. This phenomenon is commonly described as \textbf{confounding due to population stratification}. However, the mechanism of this phenomenon warrants some discussion. In particular, it is often overlooked that population structure itself does not confound the genotype-phenotype relationship in the classical sense, rather there must exist some non-genetic mechanism by which population stratification affects both the phenotype and genotype \citep{barton2019population, vilhjalmsson2012nature}. \\

@@

As an example, consider a genetic study to assess genetic variants associated with lung cancer in a sample comprised of subjects from two distinct subpopulations A and B. Assume the major allele of SNP $X$ is present with higher frequency in subpopulation A compared to subpopulation B, but it has no direct effect on lung cancer. Now suppose these subpopulations are geographically segregated in a such a way that subpopulation A is exposed to good air quality, and subpopulation B to poor air quality. Consequently, a GWAS analysis of data from these subpopulations would be predisposed to find SNP $X$ significantly associated with lung cancer. This spurious association is not identified because of the presence of population stratification itself, but because that stratification is related to a causal environmental exposure. Indeed, if subpopulations A and B were not subject to different air qualities, all else being equal, SNP $X$ would not be found to be associated with the phenotype.\\

@@

The mechanism of environmental confounding is summarized in the directed acyclic graph displayed in figure \ref{fig:ps_env}. The undirected dashed line between population stratification and environmental exposure indicates that while we assume these elements are correlated, we do not assume any causal direction. Otherwise stated, we do not assume that environmental factors directly influence population-specific allele frequencies, or vice versa, only that the two may be related. In the context of our air quality example, this means that we assume air quality is not directly altering individuals' allele frequencies, nor are allele frequencies causing individuals to move to regions of better or worse air quality. Genetic relatedness and environmental effects become associated over time for complex reasons, which are beyond the scope of this paper. As we will discuss, many methods of correcting for population structure rely on this relationship and do not formally differentiate between effects of population structure and those of environment. Such methods assume that genetic relatedness serves as an accurate proxy for environmental exposures, which may or may not be the case depending on the exposure in question.

\begin{figure}[H]
\centering
\begin{tikzpicture}[%
>=latex',
circ/.style={draw, shape=circle, node distance=1cm, line width=1pt}]%Define the arrow type and style for circled nodes
\draw[dashed] (0,1) node[left] (P) {Population Stratification} -- (2,1) node[right] (E) {Environment}; 
\draw[->] (0,0) node[left] (X) {Observed SNPs} -- (2,0) node[right] (Y) {Phenotype}; %Line between X and Y
\draw[->] (P)   -- (X) ; 
\draw[->] (E)   -- (Y) ; 
% \filldraw[color=red!60, fill=none, very thick] (0,1) ellipse (4.5 and 0.35);
\end{tikzpicture}
\caption{Directed acyclic graph (DAG) of confounding due to population structure}
\label{fig:ps_env}
\end{figure}

An additional consideration in genetic studies that we will briefly mention is that of genetic background. Genetic background describes when SNPs that have a causal relationship with the phenotype of interest are not directly observed, but are correlated with SNPs that are observed. As \citep{vilhjalmsson2012nature} note, any trait that has some genetic basis will be subject to genetic background effects, just as traits that are influenced by the environment may be subject to environmental confounding. Given the importance of both genetic and environmental contributions to many complex phenotypes, both genetic background and environmental confounding are important factors in our ability to identify true causal genetic relationships between SNPs and phenotype. Although there are a number of methods in which to account for genetic background \anna{what are they?}, the methods upon which we will focus address this by conducting a genomewide multivariate analysis.

@@

Statistical methods to control for the effects of confounding due to population stratification have proliferated as computational advances have allowed for the analysis of genetic data from large cohorts. It should be noted that by attempting to reduce the effects of population structure, these methods reduce the effects of environmental confounding by blocking the casual pathway between population structure and observed SNPs. This approach thus seeks to remove the effect of environmental heterogeneity on observed SNPs indirectly.

\anna{DAG here to illustrate this?}

The genomic control (GC) method is a univariate testing procedure which modifies a test statistic by a deflation factor. This deflation factor attempts to quantify and remove the effect of population structure on the distribution of the appropriate test statistic, and is estimated using loci independent of the phenotype of interest and the loci being tested for association \citep{devlin1999genomic, bacanu2000power, wang2009testing}. However, this deflation factor has been reported to be sensitive to the nature and quantity of the loci used to estimate it \citep{hellwege2017population, marchini2004effects}. Additionally, GC does not correct effect size estimates, so resulting coefficient estimates will be unreliable after GC is applied, even if their corresponding test statistics and p-values have been appropriately adjusted \citep{hellwege2017population}.

Stratified analysis attempts to infer subject membership within a discrete number of nonoverlapping subpopulations. Subsequent analyses are conducted within each subpopulation, and the subpopulation-specific results may then be combined via meta-analysis \citep{pritchard1999use, pritchard2000association}. By partitioning the overall sample size in this manner, the structured association method is particularly vulnerable to a loss of power. Additionally, it relies on the assumption that these subpopulations are distinct and nonoveralapping. This assumption may not accurately reflect the characteristics of a particular sample, such as one consisting of admixed populations, or recent relatedness. Considered in the context of our air quality example, rather than the existence of two subpopulations with exposure to either good or bad air quality, there may also be individuals who fall between these two extremes and are characterized by continuously varying relatedness and a gradient of air quality. By clustering those individuals within in one subpopulation or another we lose information about their fine-grained, subtle relatedness.

Another approach is to create an approximation of population structure using surrogate variables, and to adjust for these as an additional model covariates. Principal component-adjustment (PC-adjustment) methods use the first several principal components of the design matrix as these surrogate variables \citep{price2006principal}. This approach may be used in univariate or multivariate frameworks, and allows for analysis on the entirety of a structured sample.

% rely on a pair-wise similarity matrix
Like PC-adjustment methods, LMMs utilize the left singular vectors of the design matrix to estimate and subsequently correct for relatedness among subjects. However, the term LMM has been used to describe a number of distinct procedures for analyzing structured genetic data which differ in their objectives and methods. There are two main objectives of LMMs used in a structured genetic context: estimating narrow-sense heritability, and estimating individual SNP effects.

@@

LMMs can also be used to estimate SNP effects, and this can be done in either a univariate or multivariate manner. Univariate approaches attempt to estimate the marginal effect of a SNP on the phenotype and assess the statistical significance of SNP-phenotype associations while controlling for population structure using a genetic similarity matrix \citep{yu2006unified, kang2010variance, kang2008efficient}. However, as with all univariate testing approaches univariate LMM implementations may still produce spurious associations and biased effect estimates in the presence of LD.

Multivariate mixed model approaches reduce these problems by assessing the relationship between a phenotype of interest and all SNPs simultaneously via penalized regression, while controlling for environmental confounding  \citep{Rakitsch2012, bhatnagar2019simultaneous}. This approach assumes there are a relatively small number of SNPs associated with the phenotype of interest that have large effect sizes. These SNPs will be included in the model as both fixed and random effects, while SNPs with smaller effect sizes are modeled only as part of the random effect in order to control for environmental confounding.

The remainder of this review will focus on the performance of multivariate Lasso-penalized methods for the analysis of continuous, normally distributed outcomes. Specifically, PC-adjustment and LMM methods, which we will refer to as PC-Lasso and LMM-Lasso, respectively. We consider two similar implementations of the LMM-Lasso method, one developed by \citep{Rakitsch2012} and a more recent implementation from \citep{bhatnagar2019simultaneous}. Subsequent references to LMM-Lasso apply to both implementations, unless otherwise stated. Where necessary, we will differentiate between these implementations as LMM-Lasso-Rakitsch and LMM-Lasso-ggmix, respectively.

In section \ref{sec:results}, we compare the performance of PC-Lasso and LMM-Lasso in terms of their abilities to accurately estimate SNP effects and in terms of their true sign rates in the presence of varying levels of relatedness and environmental confounding. Although several studies have shown that LMMs outperform PC-adjustment methods in the univariate framework \citep{wang2013analytical, kang2010variance, zhao2007arabidopsis}, to the best of our knowledge this is the only comprehensive review and comparison of these methods in a penalized multivariate framework and using a simulation scheme motivated by human genetic relatedness and which formalizes the non-genetic confounding mechanism of population stratification.\\

>>>>

Population structure is a common phenomenon in genetic data. Varying levels of relatedness are almost always present among genetic samples, even in samples of unrelated individuals and seemingly homogeneous populations. For example, European American \citep{campbell2005demonstrating}, Han Chinese \citep{xu2009genomic, chen2009genetic}, and most recently, cohorts within the UK Biobank data \citep{haworth2019apparent} have been shown to exhibit patterns of population and geographic structure despiet their seemingly similar subjects.

Population structure is broadly categorized bsed on whether it describes recent or ancient relatedness. Recent relatedness describes familial or \textbf{cryptic relatedness}. Ancient relatedness, or common ancestry, is generally characterized by large clusters of low levels of similarity. \anna{I need some sort of subjective cutoff for recent vs. ancient relatedness rather than describing them in terms of clusters... more details} The presence of multiple distinct ancestry groups in a sample is known as \textbf{population stratification}.

Pedigree-based methods may be used to explicitly model recent relatedness if familial relationships are known. However, in this review we focus on methods to account for unobserved relatedness. Throughout this paper, the terms \textbf{population structure} and \textbf{structured population} will be used to describe a sample of individuals in which population stratification, cryptic relatedness, or both are present. Population stratification in particular has been of great concern in genetic studies due to its potential to lead to spurious associations. This phenomenon is commonly described as \textbf{confounding due to population stratification or ancestry}. The mechanism of this phenomenon warrants some discussion as it is often overlooked that population structure itself does not confound the genotype-phenotype relationship in the classical sense. While population structure may directly affect observed allele frequencies, there must also exist some mechanism by which population stratification affects the phenotype \citep{barton2019population, vilhjalmsson2012nature}, namely, the environment. \\

@@

As an example, consider a genome wide association study (GWAS) to assess genetic variants associated with lung cancer in a sample comprised of subjects from two distinct subpopulations, A and B. Assume the minor allele of SNP $X$ is present with higher frequency in subpopulation A compared to subpopulation B, but has no direct effect on lung cancer. Also suppose these subpopulations are geographically segregated in a such a way that subpopulation A is exposed to good air quality, and subpopulation B to poor air quality, and that air quality does have a direct effect on lung cancer. A GWAS of data from these subpopulations would find SNP $X$ to be significantly associated with lung cancer. This spurious association is not identified because of the presence of population stratification itself, but because that stratification is related to a causal environmental exposure. Indeed, if subpopulations A and B were not subject to different air qualities, all else being equal, SNP $X$ would not be found to be associated with the phenotype.\\

@@

This mechanism of environmental confounding is summarized in the directed acyclic graph displayed in figure \ref{fig:ps_env}. The undirected dashed line between population stratification and environmental exposure indicates that while we assume these elements are correlated, we do not assume any causal direction. Otherwise stated, we do not assume that environmental factors directly influence population-specific allele frequencies, or vice versa, only that the two may be related. In the context of our air quality example, this means that we assume air quality is not directly altering individuals' allele frequencies, nor are allele frequencies causing individuals to move to regions of better or worse air quality. Genetic relatedness and environmental effects become associated over time for complex reasons, which are beyond the scope of this paper. However, as we will discuss, many methods of correcting for population structure rely on this relationship and do not formally differentiate between effects of population structure and those of environment. Such methods assume that genetic relatedness serves as an accurate proxy for environmental exposures, which may or may not be the case depending on the exposure and population in question.

\begin{figure}[H]
\centering
\begin{tikzpicture}[%
>=latex',
circ/.style={draw, shape=circle, node distance=1cm, line width=1pt}]%Define the arrow type and style for circled nodes
\draw[dashed] (0,1) node[left] (P) {Population Stratification} -- (2,1) node[right] (E) {Environment}; 
\draw[->] (0,0) node[left] (X) {Observed SNPs} -- (2,0) node[right] (Y) {Phenotype}; %Line between X and Y
\draw[->] (P)   -- (X) ; 
\draw[->] (E)   -- (Y) ; 
% \filldraw[color=red!60, fill=none, very thick] (0,1) ellipse (4.5 and 0.35);
\end{tikzpicture}
\caption{Directed acyclic graph (DAG) of confounding due to population structure.}
\label{fig:ps_env}
\end{figure}

An additional consideration in genetic studies that we will briefly mention is that of genetic background. \anna{make the following sentence better} Genetic background describes when SNPs that have a causal relationship with the phenotype of interest are not directly observed, but are correlated with SNPs that are observed. As \citep{vilhjalmsson2012nature} note, any trait that has some genetic basis will be subject to genetic background effects, just as traits that are influenced by the environment may be subject to environmental confounding. Given the importance of both genetic and environmental contributions to many complex phenotypes, both genetic background and environmental confounding are important factors in our ability to identify true causal genetic relationships between SNPs and phenotype. Although there are a number of methods in which to account for genetic background \anna{what are they?}, the methods upon which we will focus address this by conducting a genomewide multivariate analysis.

@@

Statistical methods to control for the effects of confounding due to population stratification have proliferated as computational advances have allowed for the analysis of genetic data from large cohorts. In order to reduce confounding these methods attempt to block the causal pathway between population structure and observed SNP data. This is illustrated by the blocked arrow in figure \ref{fig:ps_env_block}. These methods, in general, do not specifically address the environmental component of confounding due to population structure. Nevertheless, they may be effective if they are able to indirectly block the causal pathway between the environment and observed SNPs. Whether or not they are effective depends upon whether population structure can be accurately inferred, and if this is a good proxy for environmental effects.

\begin{figure}[H]
\centering
\begin{tikzpicture}[%
>=latex',
circ/.style={draw, shape=circle, node distance=1cm, line width=1pt}]%Define the arrow type and style for circled nodes
\draw[dashed] (0,1) node[left] (P) {Population Stratification} -- (2,1) node[right] (E) {Environment};
\draw[->] (0,0) node[left] (X) {Observed SNPs} -- (2,0) node[right] (Y) {Phenotype}; %Line between X and Y
\draw[->, red] (P) --node {|} (X) ; 
\draw[->] (E)   -- (Y) ; 
% \filldraw[color=red!60, fill=none, very thick] (0,1) ellipse (4.5 and 0.35);
\end{tikzpicture}
\caption{The causal pathway between the environment and observed SNPs may be indirectly blocked by many methods that attempt to reduce confounding due to population structure.}
\label{fig:ps_env_block}
\end{figure}

One of the first methods to address thie problem of confounding due to population structure is the genomic control (GC) method. GC is a univariate testing procedure which modifies a test statistic by a deflation factor. This deflation factor attempts to quantify and remove the effect of population structure on the distribution of the appropriate test statistic, and is estimated using loci independent of the phenotype of interest and the loci being tested for association \citep{devlin1999genomic, bacanu2000power, wang2009testing}. However, this deflation factor has been reported to be sensitive to the nature and quantity of the loci used to estimate it \citep{hellwege2017population, marchini2004effects}. Additionally, GC does not correct effect size estimates, so resulting coefficient estimates will be unreliable after GC is applied, even if their corresponding test statistics and p-values have been appropriately adjusted \citep{hellwege2017population}.

Stratified analysis attempts to infer subject membership within a discrete number of nonoverlapping subpopulations. Subsequent analyses are conducted within each subpopulation, and the subpopulation-specific results may then be combined via meta-analysis \citep{pritchard1999use, pritchard2000association}. By partitioning the overall sample size in this manner, the structured association method is particularly vulnerable to a loss of power. Additionally, it relies on the assumption that these subpopulations are distinct and nonoveralapping. This assumption may not accurately reflect the characteristics of a particular sample, such as one consisting of admixed populations, or recent relatedness. Considered in the context of our air quality example, rather than the existence of two subpopulations with exposure to either good or bad air quality, there may also be individuals who fall between these two extremes and are characterized by continuously varying relatedness and a gradient of air quality. By clustering those individuals within in one subpopulation or another we lose information about their more fine-grained, subtle relatedness.

Another approach is to create an approximation of population structure using surrogate variables, and to adjust for these as an additional model covariates. Principal component-adjustment (PC-adjustment) methods use the first several left singular vectors of a pairwise similarity matrix as these surrogate variables \citep{price2006principal}. This approach may be used in univariate or multivariate frameworks, and allows for analysis on the entirety of a structured sample.

% rely on a pair-wise similarity matrix
Like PC-adjustment methods, LMMs utilize the left singular vectors of a similarity matrix to estimate and subsequently correct for relatedness among subjects. In an LMM, however, these left singular vectors are treated as random, rather than fixed effects. Additionally, the term LMM has been used to describe a number of distinct but related procedures for analyzing structured genetic data which differ in their objectives and methods. There are two main objectives of LMMs used in a structured genetic context: estimating narrow-sense heritability, and estimating individual SNP effects.

@@

LMMs can also be used to estimate SNP effects, and this can be done in either a univariate or multivariate manner. Univariate approaches attempt to estimate the marginal effect of a SNP on the phenotype and assess the statistical significance of SNP-phenotype associations while controlling for population structure by modeling it as a random effect using a genetic similarity matrix \citep{yu2006unified, kang2010variance, kang2008efficient}. However, as with all univariate testing approaches, univariate LMM implementations are prone to spurious associations and biased effect estimates in the presence of LD.

Multivariate mixed model approaches reduce these problems by modeling the relationship between a phenotype of interest and all SNPs simultaneously via penalized regression, while controlling for dependent error structures  \citep{Rakitsch2012, bhatnagar2019simultaneous}. This approach assumes there are a relatively small number of SNPs associated with the phenotype of interest that have large effect sizes. These SNPs will be included in the model as both fixed and random effects, while SNPs with smaller effect sizes are modeled only as part of the random effect in order to control for dependent errors due to relatedness and sample structure.

The remainder of this review will focus on the performance of such multivariate Lasso-penalized methods for the analysis of continuous, normally distributed outcomes. Specifically, PC-adjustment and LMM methods, which we will refer to as PC-Lasso and LMM-Lasso, respectively. We consider two similar implementations of the LMM-Lasso method, one developed by \citep{Rakitsch2012} and a more recent implementation from \citep{bhatnagar2019simultaneous}. Subsequent references to LMM-Lasso apply to both implementations, unless otherwise stated. Where necessary, we will differentiate between these implementations by referring to them as LMM-Lasso-Rakitsch and LMM-Lasso-ggmix, respectively.

%In section \ref{sec:results}, we compare the performance of PC-Lasso and LMM-Lasso in terms of their abilities to accurately estimate SNP effects and in terms of their true sign rates in the presence of varying levels of relatedness and environmental confounding. 
Although several studies have shown that LMMs outperform PC-adjustment methods in the univariate framework \citep{wang2013analytical, kang2010variance, zhao2007arabidopsis}, to the best of our knowledge this is the only comprehensive review and comparison of these methods in a penalized multivariate framework and using a simulation scheme motivated by human genetic relatedness and the environmental mechanism of confounding due to population structure.\\
